ExpressionView - an interactive viewer for biclusters identified in gene expression data

### 2010-01-14 expressionviewnote/expressionview.tex
More ExpressionView paper updates.

\usepackage[colorlinks,citecolor=blue,linkcolor=red,urlcolor=blue]{hyperref}
\newcommand{\thefirstpage}{1} \newcommand{\thelastpage}{1}

@@

\section{Summary:}
ExpressionView is an R package that provides an interactive graphical environment to explore biclusters identified in gene expression data. A sophisticated ordering algorithm is used to present the biclusters in a visually appealing layout that provides an intuitive summary of the results. From this overview, the user can select individual biclusters and access all the biologically relevant data associated with it. The package is aimed to facilitate the collaboration between bioinformaticians and life scientists who are not familiar with the R language.

\section{Availability:} 
\href{http://www.unil.ch/cbg/ExpressionView}{R
  package with open source license, \url{http://www.unil.ch/cbg/ExpressionView}}

@@

\subsection{Order}
To present the dozens of possibly overlapping biclusters in a visually appealing form, it is necessary to reorder the rows and columns of the gene expression matrix in such a way that biclusters form contiguous rectangles. Since it is in general impossible to find such an arrangement for more than two mutually overlapping biclusters, one has to make concessions. In contrast to \cite{grothaus06}, who proposed to repeat rows and columns as necessary to achieve this goal, we prefer to optimise the arrangement within the original data by maximising the total area of the largest contiguous biclusters.

@@

To determine the efficiency of this method, we have studied a large number of perfectly orderable, but initially scrambled, test cases. We find that in almost every situation, the proposed algorithm finds an order that recovers more than 99\% of the score of the optimal solution and in most cases, it actually recovers the correct alignment. For random samples, which are more representative for actual gene expression data than orderable situations, execution time increases polynomially with the number of clusters $m$ as ${\mathcal O}(m^\alpha)$, where $\alpha \in [1.6, 2]$, almost independently of the number of elements $n$. For a given number of clusters, we find ${\mathcal O}(n^\alpha)$, with $\alpha \in [2.5, 2.7]$.

@@

\subsection{Visualisation}
The core application of this package is the Flash applet that allows the user to interactively explore the gene expression data and the biclusters. The Flash architecture has the advantage of producing a platform and browser independent applet that can be launched from any computer with even the most restrictive user rights. A screenshot is shown in Fig.~\ref{fig:workflow}. The interface is divided in two parts: On the left-hand side, the user finds the gene expression data in the common heat map form, on top of which the biclusters are overlaid. On the right-hand side, the metadata associated with the gene expression data and the results of the enrichment calculations [GO (Gene Ontology)~\citep{ashburner00} and KEGG (Kyoto Encyclopedia of Genes and Genomes)~\citep{kanehisa04}] are shown. Wherever possible, these elements are linked to the corresponding databases. The interface essentially behaves as an image viewer, allowing the user to zoom and pan around the expression data, getting instant feedback on the selected item. The currently selected region can be exported to a PDF file, helpful to prepare figures for presentations or articles, or a simple text file.\\

%\section*{Acknowledgement}

@@

\bibitem[Ashburner et~al.(2000)]{ashburner00}
Ashburner,M. \emph{et al.} (2000)
Gene Ontology: tool for the unification of biology.
\href{http://dx.doi.org/10.1038/75556}{\emph{Nature Genetics}, {\bf 25}, 25 -- 29.}

@@

\bibitem[Madeira and Oliveira(2004)]{madeira04}
Madeira,S.C. and Oliveira,A.L. (2004)
Biclustering algorithms for biological data analysis: A survey.
\href{http://dx.doi.org/10.1109/TCBB.2004.2}{\emph{IEEE/ACM Transactions on Computational Biology and Bioinformatics}, {\bf 1}, 24 -- 45.}

\bibitem[Prelic et~al.(2006)]{prelic06}
Prelic,A., \emph{et~al.} 
A systematic comparison and evaluation of biclustering methods for gene expression data 
\href{http://dx.doi.org/10.1093/bioinformatics/bt1060}{\emph{Bioinformatics}, {\bf 22}(9), 1122 -- 1129.}

\bibitem[Santamar\'ia et~al.(2008)]{santamaria08}
Santamar\'ia,R., Ther\'on,R. and Quintales,L. (2008)
BicOverlapper: A tool for bicluster visualization. 
\href{http://dx.doi.org/10.1093/bioinformatics/btn076}{\emph{Bioinformatics}, {\bf 24}, 1212 -- 1213.}

>>>>

\usepackage{url}
% \usepackage[colorlinks,citecolor=blue,linkcolor=red,urlcolor=blue]{hyperref}
\newcommand{\thefirstpage}{1} \newcommand{\thelastpage}{1}

@@

\section{Summary:}
ExpressionView is an R package that provides an interactive graphical
environment to explore biclusters identified in gene expression
data. A sophisticated ordering algorithm is used to present the
biclusters in a visually appealing layout that provides an intuitive
summary of the results. From this overview, the user can select
individual biclusters and access all the biologically relevant data
associated with it. The package is aimed to facilitate the
collaboration between bioinformaticians and life scientists.

\section{Availability:} 
%\href{http://www.unil.ch/cbg/ExpressionView}{R 
%  package with open source license,
\url{http://www.unil.ch/cbg/ExpressionView}
%}

@@

\subsection{Order}
To present the dozens of possibly overlapping biclusters in a visually
appealing form, it is necessary to reorder the rows and columns of the
gene expression matrix in such a way that biclusters form contiguous
rectangles. Since it is in general impossible to find such an
arrangement for more than two mutually overlapping biclusters, one has
to make concessions. In contrast to \cite{grothaus06}, who proposed to
repeat rows and columns as necessary, we prefer
to optimise the arrangement within the original data by maximising the
total area of the largest contiguous biclusters.

@@

To determine the efficiency of this method, we have studied a large
number of perfectly orderable, but initially scrambled, test cases. We
find that in almost every situation, the proposed algorithm finds an
order that recovers more than 99\% of the score of the optimal
solution and in most cases, it recovers the correct
alignment. For random samples, which are more representative for
actual gene expression data than orderable situations, execution time
increases polynomially with the number of clusters $m$ as ${\mathcal
  O}(m^\alpha)$, where $\alpha \in [1.6, 2]$, almost independently of
the number of elements $n$. For a given number of clusters, we find
${\mathcal O}(n^\alpha)$, with $\alpha \in [2.5, 2.7]$.

@@

\subsection{Visualisation}
The core application of this package is the Flash applet that allows
the user to interactively explore the gene expression data and the
biclusters. The Flash architecture has the advantage of producing a
platform and browser independent applet that can be launched from any
computer with even the most restrictive user rights. A screenshot is
shown in Fig.~\ref{fig:workflow}. The interface is divided in two
parts: On the left-hand side, the user finds the gene expression data
in the common heat map form, on top of which the biclusters are
overlaid. On the right-hand side, the metadata associated with the
gene expression data and the results of the enrichment calculations
(Gene Ontology~\citep{ashburner00} and Kyoto Encyclopedia
of Genes and Genomes ~\citep{kanehisa04}] are shown. Wherever
possible, these elements are linked to the corresponding
databases. The interface essentially behaves as an image viewer,
allowing the user to zoom and pan around the expression data, getting
instant feedback on the selected item. The currently selected region
can be exported to a PDF file, or a simple text file.

\section*{Acknowledgement}

@@

\bibitem[Ashburner et~al.(2000)]{ashburner00}
Ashburner,M. \emph{et al.} (2000)
Gene Ontology: tool for the unification of biology.
\href{http://dx.doi.org/10.1038/75556}{\emph{Nature Genetics}, {\bf 25}, 25--29.}

@@

\bibitem[Madeira and Oliveira(2004)]{madeira04}
Madeira,S.C. and Oliveira,A.L. (2004)
Biclustering algorithms for biological data analysis: A survey.
\href{http://dx.doi.org/10.1109/TCBB.2004.2}{\emph{IEEE/ACM Transactions on Computational Biology and Bioinformatics}, {\bf 1}, 24--45.}

\bibitem[Prelic et~al.(2006)]{prelic06}
Prelic,A., \emph{et~al.} 
A systematic comparison and evaluation of biclustering methods for gene expression data 
\href{http://dx.doi.org/10.1093/bioinformatics/bt1060}{\emph{Bioinformatics}, {\bf 22}(9), 1122--1129.}

\bibitem[Santamar\'ia et~al.(2008)]{santamaria08}
Santamar\'ia,R., Ther\'on,R. and Quintales,L. (2008)
BicOverlapper: A tool for bicluster visualization. 
\href{http://dx.doi.org/10.1093/bioinformatics/btn076}{\emph{Bioinformatics}, {\bf 24}, 1212--1213.}
